---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A geospatial dataset of Mexican ports
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15778575.svg)](https://doi.org/10.5281/zenodo.15778575)

By: Emma Zgonena, Renato Molina & Juan Carlos Villaseñor-Derbez

**Cite as:** Zgonena et al., 2025. A geospatial dataset of Mexican ports. DOI: [10.5281/zenodo.15778574](https://zenodo.org/records/15778575#:~:text=10.5281/zenodo.15778574)

```{r map, echo = FALSE, message = FALSE, warning = FALSE}
pacman::p_load(
  here,
  rnaturalearth,
  sf,
  tidyverse
)

mex_ports <- st_read(here("data/clean/mex_ports.gpkg"),
                     quiet = TRUE)
mex <- ne_countries(country = "Mexico")

ggplot(data = mex) +
  geom_sf() +
  geom_sf(data = mex_ports,
          mapping = aes(fill = !is.na(landing_site_id)),
          color = "black",
          shape = 21,
          size = 2,
          alpha = 0.7) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.1, 0.2)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Mexican Ports",
       subtitle = "Colors indicate whether the port was matched to a CONAPESCA landing site",
       x = "Longitude",
       y = "Latitude",
       fill = "Matched")
```

## Overview

This repository contains data and code to build a geospatial dataset of the major ports of Mexico, some of which are included in CONAPESCA's landing data. It is intended to interface with other data in [mex_fisheries](https://github.com/jcvdav/mex_fisheries), such as vessel- and port-level landings information.

### How the data are built

#### 1) Extract port information and location

The final data, `mex_ports`, is a culmination of information from the files `mexian_ports.csv` and `mex_large_scale_landing_ports.csv` that are correlated through `mex_ports_dictionary.csv`.

The information originates from the pdf titled [`catastro.pdf`](/data/raw/catastro.pdf), which contains information on Mexican ports from the Mexican Secretaria de Comunicaciones y Transportes and the Departamento de Catastro, Instalaciones y Recintos. K+The following key variables were manually extracted from the PDF to create [`mexian_ports.csv`](/data/raw/mexian_ports.csv): 

- The inputs for `port_id` are labeled as 'Clv. del puerto' in `catastro.pdf`
- The inputs for `municipality_code` (renamed `municipality_id` in final `mex_ports` cleaned data) and `municipality_name` are found labeled together under 'Municipio' in `catastro.pdf`
- The inputs for `port_name` are labeled as `Nombre del Puerto` in `catastro.pdf`
- The inputs for `longitude` and `latitude` are labeled as `Longitud` and `Latutid` respectively in `catastro.pdf`

> **Note:** 155 individual ports were identified in `mexican_ports.csv`, 4 of which are on lakes and therefore are not coastline ports. 

#### 2) Identify ports with recorded fisheries landings

The information in `mex_large_scale_landing_ports.csv` is from Mexico’s fisheries production data. Code showing the extraction is included in [`01_get_large_scale_ports.R`](/scripts/01_get_large_scale_ports.R). The data do not include `port_id`, and instead use CONAPESCA's system of unique identifiers (`mex_ports_id`). We manually built a dictionary ([`mex_ports_dictionary.csv`](/data/raw/mex_ports_dictionary.csv)) that allowed us to link `port_id`s to `landinge_site_keys`.

> **Note:** 51 of the total ports in `mexican_ports.csv` can be matched to ports in `mex_large_scale_landing_ports.csv`

#### 3) Combine data

The final data, [`mex_ports`](/data/clean/), was created using the script titled [`03_combine_sources.R`](/scripts/03_combine_sources.R). 

This script cleans and retains relevant columns from the dictionary, omitting lake ports (4) that are not relevant to the set. It then combines the objects through their corresponding `port_id`. Finally, the data is standardized in name and made further accessible through the addition of tabular and geospatial file versions. 

> **Note:** The final data set has 151 ports, of which 55 can be matched to CONAPESCA's `mex_ports_id`

## Metadata

### Data files

The data are available in three formats:

- **[mex_ports.gpkg](https://github.com/mex-fisheries/mex_ports/raw/refs/heads/main/data/clean/mex_ports.gpkg)**: A geopackage in `EPSG:4326`. This is the recommended source because it is already a spatial object.
- [mex_ports.rds](https://github.com/mex-fisheries/mex_ports/raw/refs/heads/main/data/clean/mex_ports.rds): Tabular format as an R object.
- [mex_ports.csv](https://github.com/mex-fisheries/mex_ports/raw/refs/heads/main/data/clean/mex_ports.csv): Tabular format as a comma-separated values file.

### Column specifications across all are the same

- `municipality_id` or `municipality_code`: character. 5-digit code that acts as a unique identifier for each municipality. 
- `municipality_name`: character. Official name of municipality. 
- `port_id`: character. 5-digit code that acts as a unique identifier for each port. 
- `port_name`: character. Official name of port. Originally labeled under ‘Nombre del Puerto’.
- `landing_site_id`: character. 3-digit alphanumeric string that acts as a unique identifier for a what CONAPESCA considers a landing site. Note that ports not matched to a landing site id will show `NA` in this column.
- `latitude`: numeric. Latitude in decimal degrees for the location of the port.
- `longitude`: numeric. Longitude in decimal degrees for the location of the port.

## Using the data

### Build a table of landing sites an d corodinates

```{r}
#Load packages

library(tidyverse)
# Load the data using the csv file
mex_ports <- read_csv(
  file = "https://github.com/mex-fisheries/mex_ports/raw/refs/heads/main/data/clean/mex_ports.csv") # You can read the rds file using read_RDS(url("url/to/file.rds"))

# A quick glimpse at the data
glimpse(mex_ports)

# Let's get all ports that can be matched to landing site from CONAPESCA's data
mex_ports |> 
  filter(!is.na(landing_site_id)) |> 
  select(port_name, landing_site_id, longitude, latitude)
```

### Build a map

The code below is used to build the figure at the top of this document.

```{r, eval=FALSE}
# Load packages
library(rnaturalearth)
library(sf)
library(tidyverse)

# This time we read the geopackage
mex_ports_sf <- st_read("https://github.com/mex-fisheries/mex_ports/raw/refs/heads/main/data/clean/mex_ports.gpkg",
                        quiet = TRUE)

mex <- ne_countries(country = "Mexico")

ggplot(data = mex) +
  geom_sf() +
  geom_sf(data = mex_ports_sf,
          mapping = aes(fill = !is.na(landing_site_id)),
          color = "black",
          shape = 21,
          size = 2,
          alpha = 0.7) +
  theme_minimal() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.1, 0.2)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Mexican Ports",
       subtitle = "Colors indicate whether the port was matched to a CONAPESCA landing site",
       x = "Longitude",
       y = "Latitude",
       fill = "Matched")
```

